Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-16 ieee_paper.tex
section 4 partial

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers where each of them can execute 5 billion
transactions per second and has two 32-bit channels, achieving a total
bandwidth of 320GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4GB/s per core. Thus, the cores should be able to consume 504GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connect the cores and
the memory controller. Its precise bandwidth is unknown for this
precise card, but it would be coehrent that it is between 200GB/s and
250GB/s.

@@

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show that an implementation
relying on compiler vectorization does not lead to desirable
performance.) The registers in the MIC architecture are 512 bits wide
and can store 16 floats or integers. (In comparison to OpenCL or CUDA,
one can think of these registers as a warp.) The end goal is to be
able to load and format the non zero values $a,b,c,d$ and the vector
entries $p,q,r,s$ in two vectorial register to be able to apply Fused
Multiply-Add on them, which is depicted in
Figure~\ref{fig:tensor_product}. Bringing the data in the vector
register as depicted in the figure is the difficult part that we now
explain. A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each core
usign an OpenMP construct.

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of 4 lanes (sometimes called channels) of 128 bits
which are made of 4 elements of 32 bits. A swizzling operation allows
to reorder the elements within each lane. While a permutation allow to
change the orders of the lanes. Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore typecast
are necessary to convert to a vector to/from the integer type.

@@

All the operations that loads data from the memory into a register
allows to load 512 bits, so 16 floats at a time. A single non zero in
the matrix will cause 16 operations to be performed, but it only uses
4 floats from the matrix and 4 floats from the {\tt x} vector. So we
need to load at once the data for 4 non zeroes in the matrix, so that
we load 16 flaots from the matrix and 16 floats from the vectors at
once. (The number of non zeroes on a row in our application is always
a multiple of 4, if it was, we would need to add some explicit zeroes
in the data structure to pad it.)

>>>>

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers where each of them can execute 5 billion
transactions per second and has two 32-bit channels, achieving a total
bandwidth of 320GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4GB/s per core. Thus, the cores should be able to consume 504GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connect the cores and
the memory controller. Its precise bandwidth is unknown but it is
believed to be between 200GB/s and 250GB/s.

@@

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show in Section~\ref{sec:expe} that an implementation
relying on compiler vectorization does not lead to desirable
performance, which confirms the results of~\cite{Saule13-ARXIV}.) The registers in the MIC architecture are 512 bits wide
and can store 16 floats or integers. (In comparison to OpenCL or CUDA,
one can think of these registers as a warp.) The end goal is to be
able to load and format the non zero values $a,b,c,d$ and the vector
entries $p,q,r,s$ in two vector registers to be able to apply Fused
Multiply-Add on them, which is depicted in
Figure~\ref{fig:tensor_product}. Bringing the data in the vector
register as depicted in the figure is the difficult part that we now
explain. A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each core
using an OpenMP construct.

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of 4 lanes (sometimes called channels) of 128 bits
which are made of 4 elements of 32 bits. A swizzling operation allows
to reorder the elements within each lane. While a permutation allows to
change the orders of the lanes. Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore typecast
are necessary to convert to a vector to/from the integer type.

@@

All the operations that loads data from the memory into a register
allows to transfer 512 bits, so 16 floats at a time. A single non zero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we need to load at once the data for 4 non zeroes in the matrix,
representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of non zeroes on a row in our application is
always a multiple of 4, if it was, we would need to add some explicit
zeroes in the data structure to pad it.)
